State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources

### 2020-09-17 analysis/paper/paper.Rmd
Implemented some suggestions from John Hargrove, after discussing with Ryan Kinzer

---
title: "State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [BM]
      correspondence: true
  - Ryan N. Kinzer:
      email: ryank@nezperce.org
      institute: [NezPerce]
      correspondence: false
  - Michael W. Ackerman:
      email: mike.ackerman@merck.com
      institute: [BM]
      correspondence: false
institute:
  - BM: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
  - NezPerce: Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
- AckermanLibrary.bib
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  Accurate estimates of salmonids passing Lower Granite Dam on the Snake River, by species and origin, are a critical input to assessing the status and trends of various populations, as well as successful management of fisheries in the Snake River basin. Here we describe a state-space model that estimates such escapement past a dam by using window counts, passive integrated transponder (PIT) tag observations and data from an adult fish trap, accounting for issues such as nighttime passage, fallback and re-ascension, potential observation error at the window and uncertainty in the adult trap rate. We tested this approach using a simulation framework that mimicked several levels of observation error, differences between nighttime passage and re-ascension rates and the possibility of the adult trap being closed for some period of time. Our results demonstrate that the model produced unbiased estimates across all tested scenarios. We also applied this model to data from the Lower Granite dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring-summer run Chinook Salmon and steelhead from spawn years 2010-2019. 
keywords: |
  escapement; Chinook; steelhead; salmon; Lower Granite Dam; state-space model; Bayesian model
highlights: |
  These are the highlights. 
---

@@

library(tidyverse)
library(lubridate)
library(msm)
library(HDInterval)
library(ggpubr)
library(knitr)
library(kableExtra)

@@

# set default theme for ggplot2
theme_set(theme_bw())

@@

Fish escapement often refers to the number of adults that survive juvenile and subadult rearing, escape harvest and return to their natal habitat to potentially spawn (e.g. @Bue1998). For anadromous fishes, escapement is often estimated at a fixed location in a river system prior to fish reaching their spawning area. Escapement estimates facilitate effective fisheries management, particularly estimates of escapement for component groups (e.g., by stock, population, age, origin [wild, hatchery]; @hess2014monitoring; @Steinhorst2017; @Camacho2018) which provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations. Accurate escapement estimates are increasingly important for depleted populations as they inform status metrics [@mcelhany2000viable] and facilitate assessments of population viability, extinction risk [@nwfsc2015; @williams2016viability] and climate change vulnerability [@crozier2019climate].

@@

<!-- John Hargrove says: You mention in the abstract uncertainty in the trap rate as a source of error. This may be a good place to identify what error is and how it likely affects estimates of escapement.

I don't know that it's appropriate to bring that up here, since the trap rate has not been previously used to estimate escapement at all, and has been irrelevant for previous analyses.-->

@@

The second source of data came from a sample of fish collected in the adult trap as they migrated past Lower Granite Dam [@Ogden2016b]. The trap, also located within the adult fish ladder and upstream of the observation window, provided biological data (e.g., origin [wild, hatchery], genetic stock, length, age, sex) for captured adults that allowed decomposition of the escapement into specific groups (e.g., @Camacho2018, @Steinhorst2017). The trap was operational for 24 hours per day and randomly sampled the run by opening four times per hour for a length of time determined by a set daily trapping rate. The trap rate was determined by a committee of collaborating management agencies with a goal of capturing a target number of wild fish while also balancing fish handling concerns. Trap sample rates were typically 10-25%, but fluctuated throughout the season due to high water temperatures, decreased flows, trap malfunctions and/or closures, fish handling logistics, in-season forecast adjustments, etc.

<!-- John Hargrove: Are there many more items to list – if not, then consider writing them out rather than use etc. -->

@@

<!-- John Hargrove: While I think this is a noteworthy application of STADEM, as written there is no justification of why you did this or how these values compare to other estimates made at LGR. In other words, were you interested in trends in abundance? Estimated coefficients of variation? Performance relative to another estimation procedure?

We could do a comparison with the total window counts, or a comparison with SCOBI results, but I don't know that either is terribly relevant. SCOBI has been replaced by STADEM, I think. Comparing window counts with STADEM estimates of total unique fish, we could summarise the RMSE (~ 5000 for Chinook, ~10,000 for steelhead) or the normalized RMSE (~ 6% for Chinook, ! 7% for steelhead). Seem worth doing? Either that, or just explicitly state that we're doing a case study of STADEM at Lower Granite? -->

@@

@@

Weekly estimates of total escapement over Lower Granite Dam tracked the window counts and trap estimates (One example: Figure \@ref(fig:jags-fit-fig)). STADEM point estimates were often between estimates based on window counts and those based on the number of fish caught in the adult trap. However, for weeks when very few fish were caught in the trap or there was more uncertainty about the trap rate, STADEM estimates tracked the window counts more closely, as seen in the second week of July 2014, in Figure \@ref(fig:jags-fit-fig). That year also shows the utility of STADEM in dealing with missing data, as the trap was shut down for several weeks in July and August.

@@

I think showing 10 time series plots is unnecessary (Table 3 shows how estimates match up with window counts). I think 2014 is a nice illustration of general agreement, with some examples of how STADEM doesn't always track the trap estimate, and how it deals with missing data. Do we need to make that more clear, or leave it as is?-->

@@

Combining data from the adult fish trap with live and video window counts provides several benefits. First, it allows us to model observer error in the window counts, which is typically unknown. If estimates rely on window counts alone, quantifying observer error is impossible. Capturing and accounting for known sources of error is prudent to minimize management decision risk. Second, by incorporating both sources of information in a state-space framework, STADEM incorporates missing data at either the observation window or adult trap seamlessly. At Lower Granite Dam, the adult trap has been closed for brief or extended periods of time (i.e., days, weeks) intermittently over the past several years, often during peak run times. Trap closures are typically associated with elevated water temperatures resulting in potential fish handling stress and/or trap malfunctions. Given predicted Pacific Northwest climate change scenarios [@Zhang2019] trap closures from high water temperatures may become more commonplace in the future, amplifying the need for a modeling framework that accounts for periods of missing data while still capturing estimate uncertainty. Additionally, having a framework in place that accounts for missing periods of data will allow for increased logistic flexibility if, for example, maintenance or construction is needed at the observational window or adult trap.

<!-- John Hargrove on the above paragraph, specifically the sentence about the LGD adult trap having been closed over the last decade: Do you have this data for all spawn years you presented? If so, quantitatively state how common this happens. I believe these data may be presented in Ogden’s reports. Alternatively, Camacho’s reports typically list out open and close dates as well as trap rates for different spawn years. -->

@@

Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate population escapement of spring/summer Chinook Salmon and steelhead past Lower Granite Dam, and returning to tributary or population specific spawning areas [@IPTDSW2020; @kinzer2020report]. Estimates of species and origin specific escapement at Lower Granite Dam, including known uncertainty, are available to further parse into sex- or age-structured escapement estimates (e.g., @Camacho2018, @Schrader2013) that are important for fisheries management and productivity monitoring of wild populations. As an example, STADEM is being applied at Lower Granite Dam to estimate the total unique wild fish migrating past the dam. Estimates of fish passing the dam are then combined with estimated movement or transition probabilities based on PIT tag observations at instream PIT tag detection systems throughout the Snake River basin, similar to @waterhouse2020bayesian, to estimate escapement to Snake River populations and locations throughout the basin [@orme2018population]. Combined, escapement estimates from STADEM and movement probability estimates provide abundance estimates to given tributaries or populations. With sex and age data collected at the adult fish trap [@Hargrove2019], this approach provides necessary information to evaluate productivity and population viability for select Snake River Chinook Salmon and steelhead groups [@IPTDSW2020].

@@

I quickly pulled together the estimates from IDFG's latest report, 2010-2018 wild escapement. Their estimates of wild Chinook are about 4% higher than STADEM, and about 3% lower for wild steelhead. So far we've avoided explicitly comparing our results with existing methods in an effort to avoid confrontation. Maybe that's moot if IDFG has started adopting STADEM results, but looks like as of March 2019 they had not. Thoughts?-->

@@

```{r lgr-sim-tab}
crossing(`Trap rate` = c('0.15', 
                         "0.15 and 0.00 3 weeks"),
         `Fallback & Re-ascension rate` = c(0.06,
                             0.10),
         # `Re-ascension rate` = 1,
         `Window count error` = factor(c("No Error",
                                         "5% Error",
                                         "10% Error"),
                                       levels = c("No Error",
                                                  "5% Error",
                                                  "10% Error"))) %>%
  mutate(`Nighttime passage rate` = if_else(`Fallback & Re-ascension rate` == 0.06,
                                            0.06,
                                            0.05)) %>%
  mutate(Scenario = paste(rep(c('Baseline',
                                "N-R",
                                "N-R trap down",
                                "Trap down"),
                              each = 3)),
         Scenario = if_else(`Window count error` == "5% Error",
                            paste(Scenario, "Err L"),
                            if_else(`Window count error` == "10% Error",
                                    paste(Scenario, "Err H"),
                                    Scenario))) %>%
  select(Scenario, 
         `Trap rate`:`Fallback & Re-ascension rate`,
         `Nighttime passage rate`,
         `Window count error`) %>%
  kable(digits = 2,
        booktabs = T,
        linesep = "",
        caption = "Summary of simulation scenarios including varying adult trapping, fallback and re-ascension rates, nighttime passage, and window count error rates use to evaluate the performance of STADEM.") %>%
  kable_styling(latex_options = c('scale_down',
                                  "hold_position"))

@@

sum_tab %>%
  kable(digits = c(0, 3, 3, 0, 3),
        booktabs = T,
        linesep = "",
        caption = "Summary statistics, including relative bias, mean coefficient of variation (CV), root mean squared error (RMSE) and 95 credible interval coverage for results from each of the twelve simulation scenarios.") %>%
  kable_styling(latex_options = c('scale_down',
                                  "hold_position"))

@@

```{r stadem-examp-fig, fig.cap = "Schematic of how the STADEM model works. (A) shows the posterior of the estimate of fish crossing the dam while the window is open (dashed line shows observed window counts). That estimate is divided by the nighttime passage rate (B). The total fish is then discounted by the re-ascension rate to estimate unique fish (C). Those unique fish are then multiplied by the proportions by origin (D), to estimate unique fish by origin."}

arrow_length = unit(0.15, 'inches')
arrow_height = 1.05
my_xlim = c(-4, 7.5)
my_ylim = c(0.06, 1.1)

p1 = ggplot(data.frame(x = c(-5, 8)),
            aes(x)) +
  stat_function(fun = dnorm,
                args = list(mean = 4, sd = 0.4),
                aes(color = 'Daytime Fish'),
                lwd = 2) +
  stat_function(fun = dnorm,
                args = list(mean = 6, sd = 0.55),
                aes(color = 'Total Fish'),
                geom = 'blank') +
  stat_function(fun = dnorm,
                args = list(mean = 5, sd = 0.6),
                aes(color = 'Unique Fish'),
                geom = 'blank') +
  stat_function(fun = dnorm,
                args = list(mean = -1, sd = 0.75),
                aes(color = 'Unique Wild Fish'),
                geom = 'blank') +
  stat_function(fun = dnorm,
                args = list(mean = -1, sd = 0.75),
                aes(color = 'Unique Hatchery Fish'),
                geom = 'blank') +
  stat_function(fun = dnorm,
                args = list(mean = -1, sd = 0.75),
                aes(color = 'Unique HNC Fish'),
                geom = 'blank') +
  # scale_color_brewer(palette = 'Set1') +
  scale_color_brewer(palette = "Spectral") +
  # scale_color_viridis_d() +
  # scale_color_grey(start = 0.9,
  #                  end = 0.3) +
  geom_vline(xintercept = 4,
             linetype = 2) +
  coord_cartesian(xlim = my_xlim,
                  ylim = my_ylim) +
  theme_classic() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        legend.position = 'bottom') +
  labs(x = 'Escapement',
       y = 'Density',
       color = 'Fish Type',
       title = 'Daytime Fish')

p2 = p1 +
  stat_function(fun = dnorm,
                args = list(mean = 6, sd = 0.55),
                aes(color = 'Total Fish'),
                lwd = 2) +
  labs(title = 'Nighttime Passage')

@@

p3 = p2 +
  stat_function(fun = dnorm,
                args = list(mean = 5, sd = 0.6),
                aes(color = 'Unique Fish'),
                lwd = 2) +
  labs(title = 'Re-ascension')

@@

p4 = p3 +
  stat_function(fun = dnorm,
                args = list(mean = -1, sd = 0.85),
                aes(color = 'Unique Wild Fish'),
                lwd = 2) +
  stat_function(fun = dnorm,
                args = list(mean = 1, sd = 0.75),
                aes(color = 'Unique Hatchery Fish'),
                lwd = 2) +
  stat_function(fun = dnorm,
                args = list(mean = -2, sd = 1.25),
                aes(color = 'Unique HNC Fish'),
                lwd = 2) +
  geom_segment(x = 5,
               y = arrow_height,
               xend = -1,
               yend = arrow_height,
               arrow = arrow(length = arrow_length)) +
  labs(title = 'Origin Proportion')

examp_p = ggarrange(p1, p2a, p3a, p4,
                    labels = 'AUTO',
                    ncol = 2,
                    nrow = 2,
                    common.legend = T,
                    legend = 'bottom')

@@

rel_bias_p = res_long_df %>%
  filter(grepl('Unique', Variable),
         Model == 'STADEM') %>%
  mutate(Variable = fct_recode(Variable,
                               Wild = 'Unique.Wild.Fish',
                               Hatchery = 'Unique.Hatch.Fish',
                               HNC = 'Unique.HNC.Fish')) %>%
  mutate(grp = paste(Variable, Model, sep = ' / ')) %>%
  filter(Variable == "Wild") %>%
  ggplot(aes(x = Scenario,
             y = rel_bias,
             fill = grp)) +
  # stat_summary(fun.data = hdi_box,
  #              geom = 'boxplot') +
  # stat_summary(fun = hdi_out,
  #              geom = 'point',
  #              color = 'gray') +
  geom_boxplot() +
  geom_hline(yintercept = 0,
             linetype = 2) +
  labs(y = 'Relative Bias',
       fill = 'Fish / Model') +
  facet_wrap(~ Variable) +
  theme(legend.position = 'none',
        axis.text.x = element_text(angle=45, hjust = 1))

@@

```{r jags-fit-fig, fig.cap = "Time-series plot showing estimates of total escapement for Chinook in 2014, including raw window counts, window counts adjusted for nighttime passage, trap estimates and STADEM estimates. The gray ribbon represents the 95% credible interval for STADEM estimates."}

load('../data/derived_data/STADEM_results/LGR_STADEM_Chinook_2014.rda')
plotTimeSeries(stadem_mod,
               weeklyData = stadem_list$weeklyData) +
  guides(color = guide_legend(nrow = 2),
         shape = guide_legend(nrow = 2))
```

```{r night-reasc-diff-fig, fig.cap = "Nighttime passage rate plotted against re-ascension rate, calculated from observed PIT tags for each week of spawn years 2010-2019. Colors correspond to different spawn year, while the size of each point is proportional to the window count that week. The dashed line is the 1-1 line."}

rate_comp %>%
  filter(win_cnt > 0) %>%
  filter(tot_tags > 0,
         !night_rate %in% c(1, 0),
         !reasc_rate %in% c(1, 0)) %>%
  # arrange(desc(night_rate))
  ggplot(aes(x = night_rate,
             y = reasc_rate,
             color = as.factor(Year))) +
  # scale_color_viridis_d() +
  scale_color_brewer(palette = "Paired") +
  geom_abline(linetype = 2) +
  geom_point(aes(size = win_cnt)) +
  facet_wrap(~ Species,
             scales = 'free') +
  theme(legend.position = 'bottom') +
  # scale_x_continuous(trans = "log",
  #                    breaks = scales::pretty_breaks()) +
  # scale_y_continuous(trans = "log",
  #                    breaks = scales::pretty_breaks()) +
  guides(size = guide_legend(nrow = 2)) +
  labs(x = "Nighttime Passage Rate",
       y = "Re-ascension Rate",
       color = "Spawn\nYear",
       size = "Window\nCount")
```

>>>>

---
title: "State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [BM]
      correspondence: true
  - Ryan N. Kinzer:
      email: ryank@nezperce.org
      institute: [NezPerce]
      correspondence: false
  - Michael W. Ackerman:
      email: mike.ackerman@merck.com
      institute: [BM]
      correspondence: false
institute:
  - BM: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
  - NezPerce: Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      fig_width: 7
      # fig_height: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
- AckermanLibrary.bib
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  Accurate estimates of salmonids passing Lower Granite Dam on the Snake River, by species and origin, are a critical input to assessing the status and trends of various populations, as well as successful management of fisheries in the Snake River basin. Here we describe a state-space model that estimates such escapement past a dam by using window counts, passive integrated transponder (PIT) tag observations and data from an adult fish trap, accounting for issues such as nighttime passage, fallback and re-ascension, potential observation error at the window and uncertainty in the adult trap rate. We tested this approach using a simulation framework that mimicked several levels of observation error, differences between nighttime passage and re-ascension rates and the possibility of the adult trap being closed for some period of time. Our results demonstrate that the model produced unbiased estimates across all tested scenarios. We also applied this model to data from the Lower Granite dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring-summer run Chinook Salmon and steelhead from spawn years 2010-2019. 
keywords: |
  escapement; Chinook; steelhead; salmon; Lower Granite Dam; state-space model; Bayesian model
highlights: |
  These are the highlights. 
---

@@

library(tidyverse)
library(lubridate)
# library(msm)
# library(HDInterval)
library(magrittr)
library(ggpubr)
library(knitr)
library(kableExtra)

@@

# set default theme for ggplot2
# theme_set(theme_pubr(base_family = "Times New Roman",
#                      base_size = 12))
theme_set(theme_pubr())

@@

Fish escapement often refers to the number of adults that survive juvenile and subadult rearing, escape harvest and return to their natal habitat to potentially spawn [e.g. @Bue1998]. For anadromous fishes, escapement is often estimated at a fixed location in a river system prior to fish reaching their spawning area. Escapement estimates facilitate effective fisheries management, particularly estimates of escapement for component groups [e.g., by stock, population, age, origin [wild, hatchery], @hess2014monitoring; @Steinhorst2017; @Camacho2018] which provide valuable information that fisheries managers use to achieve sustainable harvest, while protecting small and vulnerable populations. Accurate escapement estimates are increasingly important for depleted populations as they inform status metrics [@mcelhany2000viable] and facilitate assessments of population viability, extinction risk [@nwfsc2015; @williams2016viability] and climate change vulnerability [@crozier2019climate].

@@

@@

The second source of data came from a sample of fish collected in the adult trap as they migrated past Lower Granite Dam [@Ogden2016b]. The trap, also located within the adult fish ladder and upstream of the observation window, provided biological data (e.g., origin [wild, hatchery], genetic stock, length, age, sex) for captured adults that allowed decomposition of the escapement into specific groups [e.g., @Camacho2018; @Steinhorst2017]. The trap was operational for 24 hours per day and randomly sampled the run by opening four times per hour for a length of time determined by a set daily trapping rate. The trap rate was determined by a committee of collaborating management agencies with a goal of capturing a target number of wild fish while also balancing fish handling concerns. Trap sample rates were typically 10-25%, but fluctuated throughout the season due to, for example, high water temperatures, decreased flows, trap malfunctions and/or closures, fish handling logistics, in-season forecast adjustments, or adjustments due to shifting species composition throughout the year.

@@

@@

All data used in this manuscript, as well as associated code for both the simulations and the Lower Granite appllication, can be found at [https://www.github.com/KevinSee/ManuscriptSTADEM]([https://www.github.com/KevinSee/ManuscriptSTADEM]).

@@

Weekly estimates of total escapement over Lower Granite Dam tracked the window counts and trap estimates (One example: Figure \@ref(fig:jags-fit-fig). Similar plots for all model runs are available on the manuscript GitHub page: [https://www.github.com/KevinSee/ManuscriptSTADEM]([https://www.github.com/KevinSee/ManuscriptSTADEM])). STADEM point estimates were often between estimates based on window counts and those based on the number of fish caught in the adult trap. However, for weeks when very few fish were caught in the trap or there was more uncertainty about the trap rate, STADEM estimates tracked the window counts more closely, as seen in the second week of July 2014, in Figure \@ref(fig:jags-fit-fig). That year also shows the utility of STADEM in dealing with missing data, as the trap was shut down for several weeks in July and August. The model's uncertainty is always smaller than the uncertainty from the trap estimates alone, whereas the window counts alone provide no estimate of uncertainty.

@@

I think showing 10 time series plots is unnecessary (Table 3 shows how estimates match up with window counts). I think 2014 is a nice illustration of general agreement, with some examples of how STADEM doesn't always track the trap estimate, and how it deals with missing data. Do we need to make that more clear, or leave it as is?

After experimenting with 4 years, Ryan and Kevin decided even 4 was too busy. Let's stick with one, and note that the rest are on the github repo-->

@@

Combining data from the adult fish trap with live and video window counts provides several benefits. First, it allows us to model observer error in the window counts, which is typically unknown. If estimates rely on window counts alone, quantifying observer error is impossible. Capturing and accounting for known sources of error is prudent to minimize management decision risk. Second, by incorporating both sources of information in a state-space framework, STADEM incorporates missing data at either the observation window or adult trap seamlessly. At Lower Granite Dam, the adult trap has been closed for brief or extended periods of time (i.e., days, weeks) intermittently in five of the last ten years, often during peak run times [@USACE2010; @USACE2011; @USACE2012; @USACE2013; @Ogden2016; @USACE2015; @USACE2016; @USACE2017; @USACE2018; @USACE2019]. Trap closures are typically associated with elevated water temperatures resulting in potential fish handling stress and/or trap malfunctions [@Ogden2016b]. Given predicted Pacific Northwest climate change scenarios [@Zhang2019] trap closures from high water temperatures may become more commonplace in the future, amplifying the need for a modeling framework that accounts for periods of missing data while still capturing estimate uncertainty. Additionally, having a framework in place that accounts for missing periods of data will allow for increased logistic flexibility if, for example, maintenance or construction is needed at the observational window or adult trap.

@@

Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate population escapement of spring/summer Chinook Salmon and steelhead past Lower Granite Dam, and returning to tributary or population specific spawning areas [@IPTDSW2020; @kinzer2020report]. Estimates of species and origin specific escapement at Lower Granite Dam, including known uncertainty, are available to further parse into sex- or age-structured escapement estimates [e.g., @Camacho2018; @Schrader2013] that are important for fisheries management and productivity monitoring of wild populations. As an example, STADEM is being applied at Lower Granite Dam to estimate the total unique wild fish migrating past the dam. Estimates of fish passing the dam are then combined with estimated movement or transition probabilities based on PIT tag observations at instream PIT tag detection systems throughout the Snake River basin, similar to @waterhouse2020bayesian, to estimate escapement to Snake River populations and locations throughout the basin [@orme2018population]. Combined, escapement estimates from STADEM and movement probability estimates provide abundance estimates to given tributaries or populations. With sex and age data collected at the adult fish trap [@Hargrove2019], this approach provides necessary information to evaluate productivity and population viability for select Snake River Chinook Salmon and steelhead groups [@IPTDSW2020].

@@

I quickly pulled together the estimates from IDFG's latest report, 2010-2018 wild escapement. Their estimates of wild Chinook are about 4% higher than STADEM, and about 3% lower for wild steelhead. So far we've avoided explicitly comparing our results with existing methods in an effort to avoid confrontation. Maybe that's moot if IDFG has started adopting STADEM results, but looks like as of March 2019 they had not.

Ryan and Kevin decided to not go down this path. One quick comparison is total escapement vs. window counts; which shows up in one of our tables, and we discuss it briefly in the results. -->

@@

```{r lgr-sim-tab}
crossing(`Trap rate` = c('0.15', 
                         "0.15 and 0.00 3 weeks"),
         `Fallback & Re-ascension rate` = c(0.06,
                             0.10),
         # `Re-ascension rate` = 1,
         `Window count error` = factor(c("No Error",
                                         "5% Error",
                                         "10% Error"),
                                       levels = c("No Error",
                                                  "5% Error",
                                                  "10% Error"))) %>%
  mutate(`Nighttime passage rate` = if_else(`Fallback & Re-ascension rate` == 0.06,
                                            0.06,
                                            0.05)) %>%
  mutate(Scenario = paste(rep(c('Baseline',
                                "N-R",
                                "N-R trap down",
                                "Trap down"),
                              each = 3)),
         Scenario = if_else(`Window count error` == "5% Error",
                            paste(Scenario, "Err L"),
                            if_else(`Window count error` == "10% Error",
                                    paste(Scenario, "Err H"),
                                    Scenario))) %>%
  select(Scenario, 
         `Trap rate`:`Fallback & Re-ascension rate`,
         `Nighttime passage rate`,
         `Window count error`) %>%
  kable(digits = 2,
        booktabs = T,
        linesep = "",
        caption = "Summary of simulation scenarios including varying adult trapping, fallback and re-ascension rates, nighttime passage, and window count error rates use to evaluate the performance of STADEM.") %>%
  kable_styling(latex_options = c('scale_down',
                                  "hold_position")) %>%
  column_spec(column = 2,
              width = "1in") %>%
  column_spec(column = c(3, 4, 5),
              width = "1.2in")

@@

sum_tab %>%
  kable(digits = c(0, 3, 3, 0, 3),
        booktabs = T,
        linesep = "",
        caption = "Summary statistics, including relative bias, mean coefficient of variation (CV), root mean squared error (RMSE) and 95\\% credible interval coverage for results from each of the twelve simulation scenarios.") %>%
  kable_styling(latex_options = c('scale_down',
                                  "hold_position"))

@@

```{r stadem-examp-fig, fig.width=3.5, fig.height=7, fig.cap = "Schematic of how the STADEM model works. (A) shows the posterior of the estimate of fish crossing the dam while the window is open (dashed line shows observed window counts). That estimate is divided by the nighttime passage rate (B). The total fish is then discounted by the re-ascension rate to estimate unique fish (C). Those unique fish are then multiplied by the proportions by origin (D), to estimate unique fish by origin."}

arrow_length = unit(0.15, 'inches')
arrow_height = 1.05
my_xlim = c(-4, 7.5)
my_ylim = c(0.08, 1.2)

p1 = ggplot(data.frame(x = c(-5, 8)),
            aes(x)) +
  stat_function(fun = dnorm,
                args = list(mean = 4, sd = 0.4),
                aes(color = 'Daytime Fish'),
                lwd = 2) +
  stat_function(fun = dnorm,
                args = list(mean = 6, sd = 0.55),
                aes(color = 'Total Fish'),
                geom = 'blank') +
  stat_function(fun = dnorm,
                args = list(mean = 5, sd = 0.6),
                aes(color = 'Unique Fish'),
                geom = 'blank') +
  stat_function(fun = dnorm,
                args = list(mean = -1, sd = 0.75),
                aes(color = 'Unique Wild Fish'),
                geom = 'blank') +
  stat_function(fun = dnorm,
                args = list(mean = -1, sd = 0.75),
                aes(color = 'Unique Hatchery Fish'),
                geom = 'blank') +
  stat_function(fun = dnorm,
                args = list(mean = -1, sd = 0.75),
                aes(color = 'Unique HNC Fish'),
                geom = 'blank') +
  # scale_color_brewer(palette = 'Set1') +
  # scale_color_viridis_d() +
  scale_color_brewer(palette = "Spectral") +
  # scale_color_grey(start = 0.9,
  #                  end = 0.3) +
  geom_vline(xintercept = 4,
             linetype = 2) +
  coord_cartesian(xlim = my_xlim,
                  ylim = my_ylim) +
  # theme_classic() +
  theme_pubr(legend = "bottom") +
  guides(color = guide_legend(nrow = 3,
                              title.position = 'top')) +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title.y = element_blank()) +
  labs(x = 'Escapement',
       color = 'Fish Type',
       title = '   Daytime Fish')

p2 = p1 +
  stat_function(fun = dnorm,
                args = list(mean = 6, sd = 0.55),
                aes(color = 'Total Fish'),
                lwd = 2) +
  labs(title = '   Nighttime Passage')

@@

p3 = p2 +
  stat_function(fun = dnorm,
                args = list(mean = 5, sd = 0.6),
                aes(color = 'Unique Fish'),
                lwd = 2) +
  labs(title = '   Re-ascension')

@@

p4 = p3 +
  stat_function(fun = dnorm,
                args = list(mean = -1, sd = 0.85),
                aes(color = 'Unique Wild Fish'),
                lwd = 2) +
  stat_function(fun = dnorm,
                args = list(mean = 1, sd = 0.75),
                aes(color = 'Unique Hatchery Fish'),
                lwd = 2) +
  stat_function(fun = dnorm,
                args = list(mean = -2, sd = 1.25),
                aes(color = 'Unique HNC Fish'),
                lwd = 2) +
  geom_segment(x = 5,
               y = arrow_height,
               xend = -1,
               yend = arrow_height,
               arrow = arrow(length = arrow_length)) +
  labs(title = '   Origin Proportion')

examp_p = ggarrange(p1, p2a, p3a, p4,
                    labels = 'AUTO',
                    ncol = 1,
                    nrow = 4,
                    common.legend = T,
                    legend = 'bottom')

@@

rel_bias_p = res_long_df %>%
  filter(grepl('Unique', Variable),
         Model == 'STADEM') %>%
  mutate(Variable = fct_recode(Variable,
                               Wild = 'Unique.Wild.Fish',
                               Hatchery = 'Unique.Hatch.Fish',
                               HNC = 'Unique.HNC.Fish')) %>%
  mutate(grp = paste(Variable, Model, sep = ' / ')) %>%
  filter(Variable == "Wild") %>%
  ggplot(aes(x = rel_bias,
             y = fct_rev(Scenario),
             fill = grp)) +
  # stat_summary(fun.data = hdi_box,
  #              geom = 'boxplot') +
  # stat_summary(fun = hdi_out,
  #              geom = 'point',
  #              color = 'gray') +
  geom_boxplot(fill = 'gray80') +
  geom_vline(xintercept = 0,
             linetype = 2) +
  # facet_wrap(~ Variable) +
  theme_pubr(legend = "none") +
    labs(x = 'Relative Bias',
       y = "Scenario",
       fill = 'Fish / Model') +
  coord_fixed(0.03)

@@

```{r jags-fit-fig, fig.cap = "Time-series plot showing estimates of total escapement for Chinook in 2014, including window counts, trap estimates and STADEM estimates. The gray ribbon represents the 95% credible intervals for STADEM estimates, while the red ribbon represents the 95% confidence intervals for the trap estimates."}

# get plotting data from particular year
plot_df = tibble(Species = "Chinook",
                 Year = "2014") %>%
  mutate(plot_title = paste(Species, Year)) %>%
  rowwise() %>%
  mutate(plot_data = map2(Species, 
                          Year,
                          .f = function(x, y) {
                            load(paste0('../data/derived_data/STADEM_results/LGR_STADEM_', Species, '_', Year, '.rda'))
                            plot_df = stadem_mod$summary[grep('^X.all', rownames(stadem_mod$summary)),] %>%
                              as_tibble(rownames = 'var') %>%
                              mutate(week = as.integer(str_extract(var, "[0-9]+")),
                                     param = str_extract_all(var, "[:alpha:]+", simplify = T)[,3],
                                     param = ifelse(param == '', 'all', param)) %>%
                              select(var, param, week, everything()) %>%
                              left_join(stadem_list$weeklyData %>%
                                          rename(week = week_num) %>%
                                          rowwise() %>%
                                          mutate(trap_lowCI = max(0, trap_est + trap_est_se * qnorm(0.025)),
                                                 trap_uppCI = trap_est + trap_est_se * qnorm(0.975)) %>%
                                          ungroup())%>%
                              mutate_at(vars(trap_est),
                                        list(~ if_else(. == Inf,
                                                       as.numeric(NA),
                                                       .))) %>%
                              select(-Species)
                            return(plot_df)
                          })) %>%
  unnest(cols = plot_data) %>%
  select(Species, Year, plot_title, everything()) %>%
  filter(param == "all")

# drop initial weeks when window counts == 0
plot_df %<>%
  group_by(Species, Year) %>%
  filter(Start_Date >= min(Start_Date[win_cnt > 0]))

# scale some variables for plotting
plot_df %<>%
    mutate_at(vars(trap_lowCI,
                 trap_uppCI,
                 `2.5%`,
                 `97.5%`,
                 win_cnt,
                 trap_est,
                 `50%`),
            list(~ . / 10000))

# set size of points
pt_size = 4
